---
title: "Exploring growth rates IWYP"
author: "A Bowerman, D Ganguly"
date: "21/05/2019"
output: pdf_document
---

### Aim 
Explore growth rates across IWYP field trial data.

### Method

Utilise measures of Normalized Vegetation Index (NDVI, Greenseeker) to estimate growth rates using a nonlinear modelling approach [Paine et al 2012](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/j.2041-210X.2011.00155.x). Also see [Fitting von Bertalanffy Growth Function](http://rpubs.com/lacs/1123) and [fishR](http://derekogle.com/fishR/examples/oldFishRVignettes/VonBertalanffy.pdf).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, warning = FALSE, message = FALSE, results = 'asis')

library(tidyverse)
library(DBI) ## functions to interface with databases
library(RMySQL) ## database implementation
library(rstudioapi)

## generate DBI COnnection Object / establish connection with database
con <- dbConnect(MySQL(),
          dbname="iwyp60_germinate_dev",
          user = "iwyp60ro",
          host = 'wheatyield.anu.edu.au')

## read tables required to pull database info
table_names <- dbListTables(con)
rq_tables <- c("phenotypedata", "phenotypes", "grouptypes", "datasets", "experiments", "germinatebase", 'groups', 'groupmembers')
tables <- lapply(FUN=dbReadTable, X=rq_tables, conn=con)
names(tables) <- rq_tables

df <- select(tables$phenotypedata, phenotype_id, germinatebase_id, dataset_id, phenotype_value) %>%
  full_join(., tables$phenotypes, by=c("phenotype_id"="id")) %>%
  full_join(., tables$germinatebase, by=c("germinatebase_id"="id")) %>%
  full_join(., tables$datasets, by=c("dataset_id"="id")) %>%
  select(-breeders_code, -breeders_name, -subtaxa_id, -puid, -colldate, -collcode, -collname, -collmissid, -othernumb, -duplsite, -duplinstname)

## disconnect from database
dbDisconnect(con)

```

### Results

The raw NDVI values can first be visualized as a jitter plot to get an idea of the existing data and for rough modelling, we can apply some polynomial models (2nd order and loess).

```{r}
library(pheatmap)
library(ggforce)

## get accession groups and members together
accs <- filter(tables$groups, grouptype_id == 3) %>%
  left_join(tables$groupmembers, by=c("id"="group_id")) %>%
  mutate(general_id = tables$germinatebase$general_identifier[match(foreign_id, tables$germinatebase$id)])

## Extract NDVI values
dat <- filter(df, sapply(strsplit(short_name,";"), function(l) l[1]) == "NDVI") %>%
  select(source_file, general_identifier, phenotype_id, germinatebase_id, experiment_id, dataset_id, short_name, phenotype_value) %>%
  mutate(trait = sapply(strsplit(short_name, ";"), function(l) l[1])) %>%
  mutate(date = sapply(strsplit(short_name, ";"), function(l) l[2])) %>%
  mutate(date = paste0(date,sapply(strsplit(source_file,"_"), function(l) l[1]))) %>%
  mutate(date = as.Date(date, tryFormats = c("%d%b%Y"))) %>%
  mutate(value = as.numeric(paste(phenotype_value))) %>%
  mutate(dataset = tables$datasets$description[match(dataset_id, tables$datasets$id)])

### plot jitter
ggplot(dat, aes(x=date, y=value)) + 
  geom_jitter() + 
  facet_wrap(~dataset) + 
  geom_smooth(method='loess') +
  geom_smooth(method="lm", formula = y ~ poly(x, 2, raw=T), color = "salmon") +
  scale_y_continuous(name = "NDVI") +
  scale_x_date(name = "Measuring date") +
  ggtitle("NDVI data") +
  theme(legend.position = "none")

```


Next, we can start to explore individual sample patterns - given the number this can be hard to do using scatterplots. Instead, let's try a heatmap.

```{r}
## make numeric matrix and heatmap
my_mat <- select(dat, germinatebase_id, date, value) %>%
  group_by(germinatebase_id, date) %>%
  summarise(avg = mean(value)) %>%
  spread(date, avg)

my_mat <- as.data.frame(my_mat)
rownames(my_mat) <- my_mat$germinatebase_id
my_mat <- as.matrix(my_mat[2:ncol(my_mat)])

# define accession annotations
my_grps <- select(accs, foreign_id, name)
my_grps <- subset(my_grps, foreign_id %in% rownames(my_mat))
rownames(my_grps) <- my_grps$foreign_id
my_grps$foreign_id <- NULL

pheatmap(mat = my_mat,
         show_rownames = F,
         main = "NDVI measures",
         annotation_row = my_grps,
         cutree_rows = 4,
         cluster_cols = FALSE,
         angle_col = 45
        )


```

The data appear to group into 4 temporal behaviours.

```{r}

```



